---
title: "cstm: Center of Spread and Total Margin"
description: |
  Using grid search to characterize ATS and total variation by NFL team.
author:
  - name: Casey Caprini
date: 2025-01-01
output:
  distill::distill_article:
    self_contained: false
    code_folding: true
---

```{r setup, include=FALSE}

library(tidyverse)
library(nflfastR)
library(nflplotR)

```


## Intro

> cstm
> Chuckling silently to myself.  Used similarly to lol, but
> when you're not actually laughing out loud.
> - Urban Dictionary

Each NFL game has a spread and point total. Over the course of a year, is a team an 'over' team, or an under team? Is there any correlation between a team's over/under and ATS results?

This is the kind of thing I heard someone describing on a recent commute. I wanted to work a grid search example to estimate something without a closed form solution, so I came up with this as a toy problem.

<b>Objective</b>: For games played in week 1 thru 17, find the point which minimizes the total euclidean distance between a team's ATS margin of victory and the over/under (i,e,. 'total margin').

### Example: 2024 Pittsburgh Steelers

I'll pull in the 2024 Pittsburgh Steelers data and plot the ATS and total margin for each of the 16 games played as of week 17.

```{r}

# Get 2024 Data
df <- nflfastR::load_pbp(2024) %>%
  filter(
    season_type == "REG" &
      (home_team == "PIT" |
         away_team == "PIT")
  ) %>%
  # Keep only relevant columns
  select(
    game_id, season, week,
    home_team, away_team,
    home_score, away_score,
    result, total,
    spread_line, total_line
  ) %>%
  # Final results are include on each row of each game.
  # For each game, keep only a single instance.
  group_by(game_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  # Calculate margins
  mutate(
    total_margin = total - total_line,
    home_ats_margin =  result - spread_line,
    away_ats_margin = -1L * home_ats_margin
  )

pit_df <-
  # Make a complete PIT record by combining PIT home games
  # with PIT away games.
  bind_rows(
    df %>%
      filter(
        home_team == "PIT"
      ) %>%
      mutate(
        team = "PIT",
        pit_ats_margin =
          home_ats_margin,
        opponent = away_team
      ) %>%
      select(
        week, team, opponent,
        pit_ats_margin, total_margin
      ),
    df %>%
      filter(
        away_team == "PIT"
      ) %>%
      mutate(
        team = "PIT",
        pit_ats_margin =
          away_ats_margin,
        opponent = home_team
      ) %>%
      select(
        week, team, opponent,
        pit_ats_margin, total_margin
      )
  ) %>%
  arrange(week)

pit_df %>%
  ggplot(
    aes(
      x = pit_ats_margin,
      y = total_margin
    )
  ) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_nfl_logos(
    aes(
      team_abbr = opponent
    ),
    width = 0.05
  ) +
  scale_x_continuous(
    breaks = seq(-60, 60, by = 6),
    minor_breaks = NULL
  ) +
  scale_y_continuous(
    breaks = seq(-60, 60, by = 6),
    minor_breaks = NULL
  ) +
  coord_fixed() +
  labs(
    x = "Steelers ATS Margin (pts)",
    y = "Game Total Margin (pts)",
    title = "2024 Steelers (Thru Week 17)",
    subtitle = "Logo = Opponent.",
    caption = "Data: nflfastR:load_pbp(2024)"
  ) +
  theme_light()

```

Next, I'll make a grid of candidate solutions to check.  For each point, I'll calculate the total euclidean distance between the candidate point and all 16 game results.

Then, I'll come to a point estimate by averaging over the the 10 points with the best performance (i.e., least total euclidean distance).

I'll use a contour plot to see patterns in total euclidean distance across the candidate grid to see if this is sensible.

```{r}

# Make a grid of values to search over.
pit_candidate_grid <-
  crossing(
    pit_ats_margin = seq(-14, 14, by = 0.5),
    total_margin = seq(-14, 14, by = 0.5)
  ) %>%
  # Metric to be minimized will be euclidean distance
  # from each candidate pair.
  mutate(
    sum_euclidean_dist = NA_real_ 
  )
  
pit_iteration_df <-
  pit_df %>%
  mutate(
    euclidean_dist = NA_real_
  )

for(i in 1:nrow(pit_candidate_grid)){
  pit_iteration_df <- pit_iteration_df %>%
    mutate(
      euclidean_dist =
        sqrt(
          (pit_ats_margin - pit_candidate_grid$pit_ats_margin[i])^2 +
            (total_margin - pit_candidate_grid$total_margin[i])^2
        )
    )
  
  pit_candidate_grid$sum_euclidean_dist[i] =
    sum(pit_iteration_df$euclidean_dist)
}

pit_estimate_df <-
  pit_candidate_grid %>%
  # Take the 10 best candidates...
  slice_min(
    order_by = sum_euclidean_dist,
    n = 10
  ) %>%
  # and average over each component
  # to develop point estimate.
  summarize(
    pit_ats_margin = mean(pit_ats_margin),
    total_margin = mean(total_margin)
  )

pit_candidate_grid %>%
  ggplot(
    aes(
      x = pit_ats_margin,
      y = total_margin
    )
  ) +
  geom_contour_filled(
    aes(
      z = sum_euclidean_dist
    ),
    binwidth = 10
  ) +
  geom_vline(xintercept = 0, color = "white") +
  geom_hline(yintercept = 0, color = "white") +
  geom_nfl_logos(
    data = pit_df,
    aes(
      team_abbr = opponent
    ),
    width = 0.05
  ) +
  geom_point(
    data = pit_candidate_grid %>%
      slice_min(
        order_by = sum_euclidean_dist,
        n = 10
      ),
    color = "red"
  ) +
  geom_point(
    data = pit_estimate_df,
    color = "white",
    shape = 2
  ) +
  scale_x_continuous(
    breaks = seq(-60, 60, by = 6),
    minor_breaks = NULL
  ) +
  scale_y_continuous(
    breaks = seq(-60, 60, by = 6),
    minor_breaks = NULL
  ) +
  coord_fixed(
    xlim = c(-14, 14),
    ylim = c(-14, 14)
  ) +
  labs(
    x = "Steelers ATS Margin (pts)",
    y = "Game Total Margin (pts)",
    title = "2024 Steelers (Thru Week 17)",
    subtitle = "Logo = Opponent. Red = 10 Best Estimates.\nWhite = Estimate (Average Over Red).",
    caption = "Data: nflfastR:load_pbp(2024)"
  ) +
  theme_light() +
  theme(legend.position = "none")
```

## A League-wide Function

Here, I write a function to do this for all teams in the league for a given season.

```{r}
# League wide for season
calculate_vegas_margins <-
  function(team_of_interest, season_of_interest = 2024L){
    
    # Load regular season data for the season of interest.
    df <- nflfastR::load_pbp(season_of_interest) %>%
      filter(
        season_type == "REG" &
          week <= 17L
      ) %>%
      select(
        game_id, season, week,
        home_team, away_team,
        home_score, away_score,
        result, total,
        spread_line, total_line
      )
    
    # Make a vector of all teams that appear in the season
    # of interest.
    teams <-
      df %>%
      pull(home_team) %>%
      unique() %>%
      sort()
    
    # Make a data frame of all final game results.
    df_results <-
      df %>%
      group_by(game_id) %>%
      slice_head(n = 1) %>%
      ungroup() %>%
      mutate(
        total_margin = total - total_line,
        home_ats_margin =  result - spread_line,
        away_ats_margin = -1L * home_ats_margin
      )
    
    home_df <-
      df_results %>%
      rename(
        team = home_team,
        opponent = away_team,
        ats_margin = home_ats_margin,
      ) %>%
      mutate(
        home_away = "home"
      ) %>%
      select(
        season, week,
        team, opponent, home_away,
        total_margin, ats_margin
      )
    
    away_df <-
      df_results %>%
      rename(
        team = away_team,
        opponent = home_team,
        ats_margin = away_ats_margin,
      ) %>%
      mutate(
        home_away = "away"
      ) %>%
      select(
        season, week,
        team, opponent, home_away,
        total_margin, ats_margin
      )
    
    long_df <-
      bind_rows(
        home_df, away_df
      ) %>%
      arrange(
        season, team, week
      )
    
    # Search grid for point that minimizes the
    # Euclidean distance from each games ATS margin
    # and over/under total margin.
    candidate_grid <-
      crossing(
        ats_margin = seq(-14, 14, by = 0.5),
        total_margin = seq(-14, 14, by = 0.5)
      ) %>%
      mutate(
        sum_euclidean_dist = NA_real_ 
      )
    
    # Stores total euclidean distance for each
    # candidate pair of ats_margin and total_margin.
    iteration_df <-
      long_df %>%
      filter(
        team == team_of_interest &
          season == season_of_interest
      ) %>%
      mutate(
        euclidean_dist = NA_real_
      )
    
    for(i in 1:nrow(candidate_grid)){
      iteration_df <- iteration_df %>%
        mutate(
          euclidean_dist =
            sqrt(
              (ats_margin - candidate_grid$ats_margin[i])^2 +
                (total_margin - candidate_grid$total_margin[i])^2
            )
        )
      
      candidate_grid$sum_euclidean_dist[i] =
        sum(iteration_df$euclidean_dist)
    }
    
    return_df <-
      candidate_grid %>%
      slice_min(
        order_by = sum_euclidean_dist,
        n = 10
      ) %>%
      summarize(
        ats_margin = mean(ats_margin),
        total_margin = mean(total_margin)
      )
    
    return(return_df)
  }

```

## 2024 NFL Results

#### Weeks 1-17

```{r}
all_teams <-
  load_pbp(2023) %>%
  pull(home_team) %>%
  unique() %>%
  sort()

all_team_results_2024 <-
  map_df(
    .x = all_teams,
    .f = calculate_vegas_margins,
    season_of_interest = 2024L
  ) %>%
  bind_cols(
    df = 
      data.frame(
        team = all_teams,
        season = 2024L
      )
  )


all_team_results_2024 %>%
  ggplot(
    aes(
      x = ats_margin,
      y = total_margin
    )
  ) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_nfl_logos(
    aes(
      team_abbr = team
    ),
    width = 0.05
  ) +
  scale_x_continuous(
    breaks = seq(-18, 18, by = 3),
    minor_breaks = NULL
  ) +
  scale_y_continuous(
    breaks = seq(-18, 18, by = 3),
    minor_breaks = NULL
  ) +
  coord_fixed() +
  labs(
    x = "Central ATS Margin (pts)",
    y = "Central Total Margin (pts)",
    title = "2024 (Thru Week 17)",
    subtitle = "Logo = Team Point Estimate.",
    caption = "Data: nflfastR:load_pbp(2024)"
  ) +
  theme_light()

all_team_results_2024 %>%
  arrange(
    desc(ats_margin)
  ) %>%
  select(season, team, everything()) %>%
  DT::datatable(
    colnames =
      list(
        "Season",
        "Team",
        "Central ATS Margin",
        "Central Total Margin"
      ),
    options = list(
      pageLength = 32
    )
  )

```

## 2023 NFL Results

#### Weeks 1-17

```{r}

all_team_results_2023 <-
  map_df(
    .x = all_teams,
    .f = calculate_vegas_margins,
    season_of_interest = 2023L
  ) %>%
  bind_cols(
    df = 
      data.frame(
        team = all_teams,
        season = 2023L
      )
  )

all_team_results_2023 %>%
  ggplot(
    aes(
      x = ats_margin,
      y = total_margin
    )
  ) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_nfl_logos(
    aes(
      team_abbr = team
    ),
    width = 0.05
  ) +
  scale_x_continuous(
    breaks = seq(-18, 18, by = 3),
    minor_breaks = NULL
  ) +
  scale_y_continuous(
    breaks = seq(-18, 18, by = 3),
    minor_breaks = NULL
  ) +
  coord_fixed() +
  labs(
    x = "Central ATS Margin (pts)",
    y = "Central Total Margin (pts)",
    title = "2023 (Thru Week 17)",
    subtitle = "Logo = Team Point Estimate.",
    caption = "Data: nflfastR:load_pbp(2023)"
  ) +
  theme_light()

all_team_results_2023 %>%
  arrange(
    desc(ats_margin)
  ) %>%
  select(season, team, everything()) %>%
  DT::datatable(
    colnames =
      list(
        "Season",
        "Team",
        "Central ATS Margin",
        "Central Total Margin"
      ),
    options = list(
      pageLength = 32
    )
  )

```

